---
title: "Analysis of MPG Based on Transmission Type"
author: "Kate Prendergast"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

In this report, we analyzed the effect of certain factors on MPG.  In particular, we 
looked at the difference between manual and automatic transmissions, and we attempted to
quantify this difference.  We used the mtcars dataset for this analysis.  We found that there
was not a significant difference in MPG between manual and automatic transmissions when
other factors were held constant.  Weight of the vehicle seemed to be a more important
factor to focus on.

Note that for the 'am' variable that designates the transmission type, 0 = automatic
and 1 = manual.

## Analysis

First we load the data and take a look at the first few rows.  For our first piece of analysis,
we fit a model that just analyzes the change in MPG against transmission type.  
```{r}
data("mtcars")
head(mtcars, 3)
fit1 <- lm(mpg ~ factor(am), data = mtcars)
summary(fit1)$coef
confint(fit1)
```

We see that there is a noticeable increase in estimated MPG (7.24) when the transmission type 
is manual (1) over automatic.  Our 95% confidence interval for the difference
(3.64 to 10.85) does not include
0, so the difference seems significant.  However, intuition tells us that other factors have a 
significant impact on MPG.  Two such factors are the weight of the car and the size of the 
engine (wt and disp, respectively).
We fit another model including these 2 variables so that we can see the change in MPG relative
to transmission type while weight and engine size are held constant.

```{r}
fit2 <- lm(mpg ~ factor(am) + wt + disp, data = mtcars)
summary(fit2)$coef
confint(fit2)
```

Now we see that the increase in MPG for manual transmission is much smaller (0.178) when 
weight and
engine size are held constant.  In fact our 95% confidence interval for the difference
(-2.86 to 3.22) includes
0, so we cannot say for sure whether manual transmission has a significant effect on MPG.
Weight seems to be a more significant driver of MPG.

Next, we compare our two models to determine if the inclusion of weight and engine size
creates a better model.

```{r}
anova(fit1, fit2)
```

Our p-value of 2.996e-07 indicates that there is a significant difference in including the
two additional variables in our second model.

Finally, we plot the density of the residuals of our second model. (See the appendix
for the plot results.)
Though it's not a perfect normal distribution, the residuals do seem to generally be
normally distributed, and the deviations may be attributable to the small size of the data set.
However, this does shed some level of doubt on our model.  Ideally, more data would be 
collected and analyzed to draw a firmer conclusion.

## Appendix

```{r}
plot(density(resid(fit2)), main = "Density of residuals of the fit2 model")
```

