---
title: "Practical Analysis of the CLM and Hypothesis Testing"
author: "Kate Prendergast"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##Part 1
###Overview
In Part 1 of this report, we will examine the distribution of means of 40 random exponentials,
using the exponential distribution (rexp in R).  This will be based on running 1000 simulations, where
each simulation creates 40 random exponentials and takes their mean.  We will then analyze the mean
and variance of our simulated data (as compared to the theoretical values) and look at the distribution.

###Simulations
We first create our 1000 simulations, taking the mean of 40 exponentials in each.  For this analysis,
we will use a lambda of 0.2 (the second argument to the rexp function).
```{r}
## set the seed for the sake of reproducibility of this report
set.seed(22)
mns = replicate(1000, mean(rexp(40, 0.2)))
hist(mns, xlab = "Mean", main = "Histogram of Means of Simulated Data")
```

We can see that the sample means are centered around 5.


###Sample Mean vs. Theoretical Mean
We now calculate the exact mean of our simulated data.
```{r}
simMean <- mean(mns)
```
This results in a sample mean of **`r simMean`**.  The expected, theoretical mean of the exponential
distribution is 1/lambda.  The lambda used in our simulation was 0.2, which results in an expected 
mean of **5**.  Hence the mean of our simulated data is very close to the expected value.

###Sample Variance vs. Theoretical Variance
Next, we calculate the variance of our simulated data.
```{r}
simVar <- var(mns)
```
This results in a sample variance of **`r simVar`**.  The expected, theoretical variance of the exponential
distribution (taking into account the n = 40 draws from the population) is (1/n) * (1/lambda^2), which
results in an expected variance of **0.625**.  Once again, the variance of our simulated data is very
close to the expected value.

###Distribution
Finally, we want to see if the distribution of our sample means is approximately normal. We first 
show a side-by-side comparison of a plot of 1000 random exponentials (using our same lambda of 0.2)
and the same histogram of our means of our simulations of 40 samples.
```{r}
par(mfrow=c(1,2))
hist(rexp(1000, 0.2), xlab = "Exponentials", main = "Distribution of Exponentials")
hist(mns, xlab = "Means", main = "Distribution of Sample Means")
```

As expected, the first plot shows the frequency of values decreasing exponentially, with the
majority of values near 5.  Our second plot also shows the means centered around 5,
however in this plot we see that our distribution of means takes on a Gaussian shape.

As an additional check, we plot the density of the simulated means in the figure below, and we have 
overlayed a normal curve.
```{r}
hist(mns, xlab = "Mean", main = "Histogram of Means of Simulated Data", prob = TRUE)
curve(dnorm(x, mean = simMean, sd = sqrt(simVar)), col = "blue", add = TRUE, lwd = 2)
```

We can see that the distribution of our sample means is indeed approximately normal.

###Summary
In summary, our analysis shows that the simulated data does behave as predicted by the CLT.
Our mean and variance approach the theoretical values, and the distribution of our sample 
means is approximately normal.

## Part 2
###Overview
In Part 2 of this report, we will examine a fairly small dataset that shows tooth
growth results from Vitamin C, using one of two delivery methods and given at three
different doses.  We will examine whether the data appears to be normally distributed,
and if so, determine if the differences in resutls appear to be statistically
significant.

###Data Summary
We first load the ToothGrowth dataset and use the stat.desc() function to perform some initial
analysis of the data.
```{r}
library(datasets)
library(pastecs)
data("ToothGrowth")
head(ToothGrowth)
stat.desc(ToothGrowth, norm = TRUE)
```

Based on the Shapiro-Wilk normality test statistic p-value of 0.11 calculated for this dataset
and an assumed significance level of 0.05, we will not reject the hypothesis that the data is 
normally distributed and can proceed with confidence estimates using a t-distribution.

We start by plotting the results by dose and delivery method.  Dose values are 0.5, 
1, and 2.  Delivery methods are OJ and VC, which stand for Orange Juice and Ascorbic
Acid, respectively.

```{r}
library(ggplot2)
qplot(factor(dose), len, data = ToothGrowth, geom = "boxplot", facets = .~supp, xlab = "Dose", ylab = "Tooth Growth Length", main = "Tooth Growth Length by Dose and Delivery Method")
```

There doesn't appear to be significant differences by delivery method, but the dose
does seem to affect the length of tooth growth.  Let's further examine this using
t-test confidence intervals.

###Testing
We start by subsetting the data by dose so that we can then compare two at a time.
We then compare the 0.5 and 1.0 doses.  We look at the confidence interval, which
will show with 95% confidence what the difference in the mean is between the two
data sets.
```{r}
dose05 <- subset(ToothGrowth, dose == 0.5)
dose1 <- subset(ToothGrowth, dose == 1.0)
dose2 <- subset(ToothGrowth, dose == 2.0)
t.test(dose05$len, dose1$len)$conf.int
```

Since 0 is not in the interval, we can say with 95% confidence that the difference
in the means between the 0.5 and 1.0 doses is significant (non-zero).

We continue the analysis by comparing the 1.0 and 2.0 dose data, as well as the 0.5
and 2.0 dose data (to cover all combinations, though logic tells us that there is a
significant difference in the means between the 0.5 and 2.0 dose data.)

```{r}
t.test(dose1$len, dose2$len)$conf.int
t.test(dose05$len, dose2$len)$conf.int
```

Once again, with both of these tests, 0 is not in the confidence interval, so the
differences in the means between the doses appears to be significant (non-zero).

