---
title: "Practical Analysis of the CLM and Hypothesis Testing"
author: "Kate Prendergast"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##Part 1
###Overview
In Part 1, we will examine the distribution of means of 40 random exponentials,
using the exponential distribution.  This will be based on running 1000 simulations, where
each one takes the mean of 40 random exponentials.  We will analyze the mean
and variance of our simulated data and look at the distribution.

###Simulations
We first create our 1000 simulations, taking the mean of 40 exponentials in each.  For this analysis,
we will use a lambda of 0.2.
```{r}
## set the seed for the sake of reproducibility of this report
set.seed(22)
mns = replicate(1000, mean(rexp(40, 0.2)))
hist(mns, xlab = "Mean", main = "Histogram of Means of Simulated Data")
```

We can see that the sample means are centered around 5.

###Sample Mean vs. Theoretical Mean
```{r echo=FALSE}
simMean <- mean(mns)
simVar <- var(mns)
```
We now calcualte the exact mean of our data, which gives us **`r simMean`**.  The theoretical mean of the exponential
distribution (1/lambda) gives an expected 
mean of **5**.  Hence the mean of our simulated data is very close to the expected value.

###Sample Variance vs. Theoretical Variance
Next, we calculate the variance of our simulated data, which is **`r simVar`**.  The theoretical variance of the exponential
distribution (with n = 40 draws from the population) is (1/n) * (1/lambda^2), which
gives us  **0.625**.  Once again, the variance of our simulated data is very
close to the expected value.

###Distribution
Finally, we see if the distribution of our sample means is approximately normal. We 
show a comparison of a plot of 1000 random exponentials (using our same lambda of 0.2)
and the same histogram of our means of our simulations of 40 samples.

```{r echo=FALSE}
par(mfrow=c(1,2))
hist(rexp(1000, 0.2), xlab = "Exponentials", main = "Distribution of Exponentials")
hist(mns, xlab = "Means", main = "Distribution of Sample Means")
```

As expected, the first plot shows the frequency of values decreasing exponentially, with the
majority of values near 5.  Our second plot also shows the means centered around 5,
however in this plot we see that our distribution of means takes on a Gaussian shape
and is normally distributed.

```{r echo=FALSE, results='hide'}
##As an additional check, we plot the density of the simulated means in the figure below, and we have 
##overlayed a normal curve.
##```{r}
##hist(mns, xlab = "Mean", main = "Histogram of Means of Simulated Data", prob = TRUE)
##curve(dnorm(x, mean = simMean, sd = sqrt(simVar)), col = "blue", add = TRUE, lwd = 2)
##```


##We can see that the distribution of our sample means is indeed approximately normal.
```

###Summary
In summary, our analysis shows that the simulated data does behave as predicted by the CLT.
Our mean and variance approach the theoretical values, and the distribution of our sample 
means is approximately normal.

## Part 2
###Overview
In Part 2, we will examine a fairly small dataset that shows tooth
growth results from Vitamin C, using one of two delivery methods and given at three
different doses.  We will examine whether the data appears to be normally distributed,
and if so, determine if the differences in resutls appear to be statistically
significant.

###Data Summary
We first load the ToothGrowth dataset and use the stat.desc() function to perform some initial
analysis of the data.  We are most interested in the Shapiro-Wilk normality test
results, shown below, but see the appendix for the full analysis.
```{r echo=FALSE}
library(datasets)
library(pastecs)
```

```{r echo=FALSE}
data("ToothGrowth")
stat.desc(ToothGrowth, norm = TRUE)["normtest.p",]
```

Based on the Shapiro-Wilk normality test statistic p-value of 0.11 calculated for this dataset
and an assumed significance level of 0.05, we will not reject the hypothesis that the data is 
normally distributed and can proceed with confidence estimates using a t-distribution.

###Testing
We start by plotting the results by dose and delivery method.  Dose values are 0.5, 
1, and 2.  Delivery methods are OJ and VC, which stand for Orange Juice and Ascorbic
Acid, respectively.

*See Figure 1 in the appendix for the accompanying boxplot.*

There doesn't appear to be significant differences by delivery method, but the dose
does seem to affect the length of tooth growth.  Let's further examine
the differences in results by dose usingt-test confidence intervals.

After subsetting the data by dose (see Appendix for code) so that we can compare two at a time, we then compare the 0.5 to 1.0 doses, 1.0 to 2.0, and 0.5 to 2.0  We look at the confidence interval for each comparison, which
will show with 95% confidence what the difference in the mean is between the two
data sets involved.
```{r echo=FALSE}
dose05 <- subset(ToothGrowth, dose == 0.5)
dose1 <- subset(ToothGrowth, dose == 1.0)
dose2 <- subset(ToothGrowth, dose == 2.0)
```

```{r}
rbind(as.numeric(t.test(dose05$len, dose1$len)$conf.int),
as.numeric(t.test(dose1$len, dose2$len)$conf.int),
as.numeric(t.test(dose05$len, dose2$len)$conf.int))
```

In all 3 cases, since 0 is not in the interval, we can say with 95% confidence that the difference in the means is significant (non-zero) and that dose does affect
the tooth growth results.

##Appendix
###Part 1
Code for calculating the sampling mean and variance.

```{r results='hide'}
simMean <- mean(mns)
simVar <- var(mns)
```

Code for creating the side-by-side comparison graphs.
```{r echo=TRUE, results='hide', eval=FALSE}
par(mfrow=c(1,2))
hist(rexp(1000, 0.2), xlab = "Exponentials", main = "Distribution of Exponentials")
hist(mns, xlab = "Means", main = "Distribution of Sample Means")
```

###Part 2
Code for loading the ToothGrowth data and performing a basic statistical analysis.
```{r}
data("ToothGrowth")
stat.desc(ToothGrowth, norm = TRUE)
```

Figure 1 - code and boxplot
```{r}
library(ggplot2)
qplot(factor(dose), len, data = ToothGrowth, geom = "boxplot", facets = .~supp,
      xlab = "Dose", ylab = "Tooth Growth Length", 
      main = "Tooth Growth Length by Dose and Delivery Method")
```

Code for subsetting tooth decay data.
```{r results='hide'}
dose05 <- subset(ToothGrowth, dose == 0.5)
dose1 <- subset(ToothGrowth, dose == 1.0)
dose2 <- subset(ToothGrowth, dose == 2.0)
```