---
title: "Hypothesis Testing - Tooth Growth Sample Data"
author: "Kate Prendergast"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

###Overview
In this report, we will examine a fairly small dataset that shows tooth
growth results from Vitamin C, using one of two delivery methods and given at three
different doses.  We will examine whether the data appears to be normally distributed,
and if so, determine if the differences in resutls appear to be statistically
significant.

###Data Summary
We first load the ToothGrowth dataset and use the stat.desc() function to perform some initial
analysis of the data.  
```{r}
library(datasets)
library(pastecs)
data("ToothGrowth")
stat.desc(ToothGrowth, norm = TRUE)
```

Based on the Shapiro-Wilk normality test statistic p-value of 0.11 calculated for this dataset
and an assumed significance level of 0.05, we will not reject the hypothesis that the data is 
normally distributed and can proceed with confidence estimates using a t-distribution.

###Testing
We start by plotting the results by dose and delivery method.  Dose values are 0.5, 
1, and 2.  Delivery methods are OJ and VC, which stand for Orange Juice and Ascorbic
Acid, respectively.

```{r}
library(ggplot2)
qplot(factor(dose), len, data = ToothGrowth, geom = "boxplot", facets = .~supp,
      xlab = "Dose", ylab = "Tooth Growth Length", 
      main = "Tooth Growth Length by Dose and Delivery Method")
```


There doesn't appear to be significant differences by delivery method, but the dose
does seem to affect the length of tooth growth.  Let's further examine
this using t-test confidence intervals.

After subsetting the data by dose so that we can compare two at a time, we then compare the 0.5 to 1.0 doses, 1.0 to 2.0, and 0.5 to 2.0  We look at the confidence interval for each comparison, which
will show with 95% confidence what the difference in the mean is between the two
data sets involved.
```{r}
## subset the data by dose
dose05 <- subset(ToothGrowth, dose == 0.5)
dose1 <- subset(ToothGrowth, dose == 1.0)
dose2 <- subset(ToothGrowth, dose == 2.0)
## perform test for each combination, binding resutls together for formatting ease
rbind(as.numeric(t.test(dose05$len, dose1$len)$conf.int),
as.numeric(t.test(dose1$len, dose2$len)$conf.int),
as.numeric(t.test(dose05$len, dose2$len)$conf.int))
```

In all 3 cases, since 0 is not in the interval, we can say with 95% confidence that the difference in the means is significant (non-zero) and that dose does affect
the tooth growth results.

Now we repeat this analysis comparing the two methods of delivery.

```{r}
## subset the data by delivery method
oj <- subset(ToothGrowth, supp == "OJ")
vc <- subset(ToothGrowth, supp == "VC")
## perform test comparing the two methods
as.numeric(t.test(oj$len, vc$len)$conf.int)
```

As suspected, the confidence interval does contain 0, so we cannot rule out the
possibility that there is no difference in the means of the data for the two
delivery methods.

###Summary
Based on our t-test confidence interval testing, there does seem to be a significant
difference in results based on the dose but not based on the delivery method.  This is
based on the 95% confidence intervals of our tests and the assumption of a 0.05
significance level on the Shapiro-Wilk normality test.