---
title: "Health and Economic Impact of Weather Events By Type"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Synopsis
In this report, we will examine data from the National Oceanic and Atmospheric Administration (NOAA)
on severe weather events.  The NOAA data tracks many data points for each severe weather event in the U.S.,
including fatalities, injuries, and damage to property and crops.  Our aim is to determine which type of events
are most harmful to population health and which have the greatest economic consequences.  From this analysis,
we found that tornados and floods are the events that have the most impact on population health and economic factors.

## Data Processing
First, we read the data in from the csv file with the NOAA data.
```{r cache=TRUE}
stormData <- read.csv("StormData.csv.bz2")
```

```{r}
dim(stormData)
```

We can see there are 902,297 observations of 37 variables.  Let's take a look at the first few rows.

```{r}
head(stormData)
```

## Results

### Population Health Impact

We first want to look at the impact on population health by event type.  For this analysis, we will combine both
injuries and fatalities when looking at the effect on health.

First, we will create a new column with the sum of fatalities and injuries.  Then we sum this field by event type
and sort the results in descending order to see the event types with the most fatalities and injuries.

```{r results='markup'}
library(dplyr)
stormData$PopHealth <- stormData$FATALITIES + stormData$INJURIES
popHealthSums <- aggregate(PopHealth~EVTYPE, data = stormData, sum)
popHealthSums <- arrange(popHealthSums, desc(PopHealth))
```
The bar chart below shows the top 5 events that are the most harmful to population health.

```{r}
barplot(head(popHealthSums, n=5)$PopHealth, names.arg = head(popHealthSums, n=5)$EVTYPE, cex.names = 0.7, xlab = "Event Type", ylab = "Total Fatalities and Injuries", main = "Event Type Impact on Population Health")
```

### Economic Impact

To analyze the events with the greatest economic impact, we will look at the property damage and crop damage of
each event.  This data is stored in 2 fields for each type of damage - a numeric amount and a character exponent.
For example, $25,000 could be stored as a numeric value of 25 and an exponent of K.  According to the NOAA documentation,
characters for exponent should either be K (thousands), M (millions), or B (billions).  Any other exponent values
will be ignored, and such rows will not be included in the analysis, as it is not clear what they signify.  We will,
however, include lowercase or uppercase values of these three exponents.

We first create a new data set that only includes the event type and damage data in order to minimize the amount of data
involved. We then add a new column to hold the total damage amount and assign it a value of zero for all rows.
```{r}
dmgData <- select(stormData, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
dmgData$TotalDamage <- 0
```

Next we add the property and crop damage to the total damage field, multiplying each by the appropriate factor of 10
based on the extension.
```{r}
## increase damage total by property damage * 1000 for K/k exponents
dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("K", "k")] <- dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("K", "k")] + (dmgData$PROPDMG[dmgData$PROPDMGEXP %in% c("K", "k")] * 1000)
## increase damage total by property damage * 1000000 for M/m exponents
dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("M", "m")] <- dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("M", "m")] + (dmgData$PROPDMG[dmgData$PROPDMGEXP %in% c("M", "m")] * 1000000)
## increase damage total by property damage * 1000000000 for B/b exponents
dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("B", "b")] <- dmgData$TotalDamage[dmgData$PROPDMGEXP %in% c("B", "b")] + (dmgData$PROPDMG[dmgData$PROPDMGEXP %in% c("B", "b")] * 1000000000)

## increase damage total by crop damage * 1000 for K/k exponents
dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("K", "k")] <- dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("K", "k")] + (dmgData$CROPDMG[dmgData$CROPDMGEXP %in% c("K", "k")] * 1000)
## increase damage total by crop damage * 1000000 for M/m exponents
dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("M", "m")] <- dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("M", "m")] + (dmgData$CROPDMG[dmgData$CROPDMGEXP %in% c("M", "m")] * 1000000)
## increase damage total by crop damage * 1000000000 for B/b exponents
dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("B", "b")] <- dmgData$TotalDamage[dmgData$CROPDMGEXP %in% c("B", "b")] + (dmgData$CROPDMG[dmgData$CROPDMGEXP %in% c("B", "b")] * 1000000000)
```

Now we have the total damage for each row in the data set.  We will sum the damage by event type and again order the 
results in descending order so that we can look at those events that have the highest economic impact.
```{r}
dmgEventSums <- aggregate(TotalDamage~EVTYPE, data = dmgData, sum)
dmgEventSums <- arrange(dmgEventSums, desc(TotalDamage))
```

The bar chart below shows the top 5 events with the greatest economic impact.

```{r}
barplot(head(dmgEventSums, n=5)$TotalDamage, names.arg = head(dmgEventSums, n=5)$EVTYPE, cex.names = 0.7, xlab = "Event Type", ylab = "Total Damage in USD", main = "Event Type Economic Impact")
```

### Summary
We can see that tornados have the biggest impact on population health and that flooding has the greatest economic
impact.  Also, these are the only 2 events that occur in the top 5 of both categories.