---
title: "Week 4 Pitch Presentation"
date: "March 8, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome

Welcome to my Shiny app pitch.  The app can be found at <https://firmbeliever.shinyapps.io/week4app/>.

## Overview

The app uses data from the 'cars' dataset from R.  The dataset contains 50 observations of the speed of
a car (in MPH) and its stopping distance (in feet).  The app shows the plotted dataset, and a linear
 model is shown by a blue line.  
 
 The user can use a slider to pick a speed between 5 and 40 MPH, and the app 
 will predict the stopping distance.  This prediction will be listed below the slider, and it 
 will also be plotted by a red dot on the blue line in the graph.

## Example

Below is an example showing the model that is fit for the dataset, as well as a sample prediction
with a new speed of 40 MPH.

```{r cars, echo = TRUE, comment=""}
model <- lm(dist ~ speed, data = cars)
predict(model, newdata = data.frame(speed = 40))[[1]]
```

## Summary

This app could be useful for educating users on the need to maintain proper distance,
particularly when traveling at higher speeds.


